Hijacking Simulators with Universal Probabilistic Programming

### 2019-07-30 paper/learning/notes.tex
updated

\paragraph{Models of decay of natural immunity}
\textbf{To complete}
  % \begin{align}
%   \mu_i &= N \\ % pop size
%   % disease dynamics
%   S_{\infty} &\sim \mathcal{N}(\mu_\infty, \sigma^{2}_\infty) \\
%   E_{crit} &\sim \mathcal{N}(\mu_{crit}, \sigma^{2}_{crit}) \\
%   S_{imm} &\sim \mathcal{N}(\mu_{imm}, \sigma^{2}_{imm}) \\
%   X_{p_{crit}} &\sim \mathcal{N}(\mu_{p_{crit}}, \sigma^{2}_{p_{crit}}) \\
%   \gamma_p &\sim \mathcal{N}(\mu_{p}, \sigma^{2}_{p}) \\
%   \rho_i & \sim \mathcal{U}(\mathbb{L}_i , \mathbb{U}_i) \\
%   % integration for parasite densities in different demographics
%   for \hspace{0.1cm} t &= 1:t_{max}: \\
%   E_a &= \frac{E_{{max}_t}}{A_{max}} \times h_{\mu_{SA}}() \\
%   X_p &= E_a + X_p \\
%   S_1 &= S_{\infty} + \frac{(1 - S_\infty) }{1 + \frac{E_a}{E_{crit}}} \\
%   S_2 &= S_{imm} + \frac{1 - S_{imm}}{1 + (\frac{X_p}{X_p_{crit}})^{\gamma_p}}\\
%   %Probability of survival function
%   S_p &= S_1 \times S_2 \\
%   \lambda &= S_p \times E_a \\
%   % Number of infections generated in unit time per individual sampled from poisson
%   h &\sim Pos(\lambda) \\
%   \texttt{if} &\hspace{0.1cm} h > 0: \\
%   \hspace{2.3cm} &n_{infect} = 1 \\
%   \text{TODO: add prevelance construction below}
%   % % end for - add prevelance construction below
%   % \mathcal{P}rev &= 
% \end{align}

@@

\section{Details}

We do this by  Replacing the rejection sampling loops with 
learnt functions from an amortized inference procedure Figure~\ref{fig:amortized_sampling}.
This then enables us to replace rejection sampling loops with amortized functions that 
are quick to evaluate, provide a functional form to the density and significantly reduce 
memory over-heads, as compared to~\cite{baydin2018efficient}, due to not having to track the 
thousands of raw sample calls $f_{a_i}$ generated in standard simulators.

@@

\begin{figure}
  \centering
  \includegraphics{rs_plots.png}
  \caption{ Simulated densities for rejection sampling block 1, \emph{blue} and rejection
  sampling block 2, \emph{orange}.}
  \label{fig:rs_simulated}
\end{figure}
However, in order todo this, we must correctly specify our variational objective.
\section{ An overview of Amortized Inference schemes}

\begin{itemize}
  \item Whilst the variational posterior we choose as our proposal has a functional form 
  and can be sampled, the functional form of the rejection sampling loops is not explicitly known,
  which leads to problems when calculating the $KL$-divergence. 
  \item There are two approaches that we can take to calculating some divergence metric. 
  \item 1) Using something like a non-parametric divergence estimator that makes use of Renyi and $L_{2}$ divergences (see \cite{basseville2013divergence} for details)
  \item 2) Use some form of kernel density to estimation to approximate the functional form of the rejection sampling 
  block, which can be used in place. From which we learn a variational approximation characterized by a trained network that we can sample from quickly. 
  \item An issue with approach 2) is that if I can approximate the density in functional form aren't there more efficient methods that we can use? (\textbf{Ask tom})
  \item Or, if assuming that the rejection sampling block can be arbitrary dimension, are we first learning a flow to represent that density and then learn a variational approximation of something that we can easily sample from. 
  \item But, this is very much the idea behind normalising flows, take a simple distribution that you can sample and then map it to the ``complicated'' density. 
  \item Stein Variational inference may be quite nice to use here~\cite{feng2017learning}, but it does make explicit assumptions in requiring the densities to be differentiable. 
  \item  For an overview of all variational methods, including the use of different divergence metrics see here ~\cite{8588399}
\end{itemize}
  \subsection{Using variants of VAEs}

@@

\bibliography{refs}
\bibliographystyle{plain}

>>>>

\paragraph{Models of decay of natural immunity}
\textbf{To complete}

@@

\section{Rejection Sampling}

Rejection samplers are heavily used when designing simulators, due to their ease of use and ability to model 
any arbitrary distribution on $\mathbb{R}^{D}$, with a sufficiently well designed rejection sampling scheme~\cite{MEYER20083408,martino2012improved,casella} and have 
their origins in 1940 - 1950's, during the development of other prominent Monte carlo techniques at Los Alamos.

In the standard setting we let $X$ be a random variable whose density is $p(x)$ that we cannot easily 
sample from, but can evaluate. Using a random variable $Z$, whose density is $g(z)$  that we can efficiently
sample from, then we can generate a sample $X$ by sampling $Z$ and accept $Z$ with probability
$p_{accept} = \frac{p(x)}{Mg(x)}$, which we can do until acceptance. 
M is constant that has a finite bound, $M \in (1, \infty)$, on the likelihood ratio $\frac{p(x)}{g(x)}$ over the 
support of $X$. $M$ must also satify the following inequality $p(x) \leq M g(x)  \forall x$. 
This implies that the support of $Z$ must also contained the support of $X$, that is
$g(x) > 0$ whenever $p(x)>0$. 
The standard algorithm is defined as follows:
\begin{algorithm}
  \small
   \caption{Rejection Sampling}
    \begin{algorithmic}\small
       \State {At iteration $i$ s.t $i \geq 1$}
       \State $X_{i} \sim g(X_{i})$ 
       \State $U_{i} \sim \mathcal{U}(0,1)$
       \If{$U_{i} \leq M \frac{f(X_i)}{g(X_{i})}$}
       \State{ accept $X_i \sim f$}
       \Else
       \State{$i \gets i + 1$, repeat}
       \EndIf
\end{algorithmic}
\end{algorithm}

As useful rejection sampling is, it has many short comings. 
In particular, to simulate proposals for certain distributions,
such as truncated normal distributions, regardless of the easy to sample from proposal
chosen the computational running time of such rejection statements can be shown to have an
expectation equal to infinity. Thus, the ability to learn and replace such rejection
sampling blocks with learnt surrogate functions, that are expressive enough to 
reciprocate the behavior of the rejection sampling block buys the user two things. 
A) It reduces the computational complexity of the simulator. 
B) As we now have a direct functional form of the density, we can correctly leverage
more computationally and sample efficient inference schemes. 
\section{Details}

We do this by  Replacing the rejection sampling blocks with 
learnt functions from an amortized inference procedure Figure~\ref{fig:amortized_sampling}.
This then enables us to replace rejection sampling loops with amortized functions that 
are quick to evaluate, provide a functional form to the density and significantly reduce 
memory over-heads, as compared to~\cite{baydin2018efficient}, due to not having to track the 
thousands of raw sample calls $f_{a_i}$ generated in standard simulators.

@@

\begin{figure}
  \centering
  \includegraphics{rs_plots.png}
  \caption{ Simulated densities for rejection sampling block 1, \emph{blue} and rejection
  sampling block 2, \emph{orange}.}
  \label{fig:rs_simulated}
\end{figure}
However, in order to do this, we must correctly specify our variational objective.

\section{ An overview of Amortized Inference schemes}
\textbf{Bradley: Todo: Tom has provided me with all the information to conduct this approach}
\begin{itemize}
  \item Whilst the variational posterior we choose as our proposal has a functional form 
  and can be sampled, the functional form of the rejection sampling loops is not explicitly known,
  which leads to problems when calculating the $KL$-divergence. 
  \item There are two approaches that we can take to calculating some divergence metric. 
  \item 1) Using something like a non-parametric divergence estimator that makes use of Renyi and $L_{2}$ divergences (see \cite{basseville2013divergence} for details)
  \item 2) Use some form of kernel density to estimation to approximate the functional form of the rejection sampling 
  block, which can be used in place. From which we learn a variational approximation characterized by a trained network that we can sample from quickly. 
  \item An issue with approach 2) is that if I can approximate the density in functional form aren't there more efficient methods that we can use? (\textbf{Ask tom})
  \item Or, if assuming that the rejection sampling block can be arbitrary dimension, are we first learning a flow to represent that density and then learn a variational approximation of something that we can easily sample from. 
  \item But, this is very much the idea behind normalising flows, take a simple distribution that you can sample and then map it to the ``complicated'' density. 
  \item Stein Variational inference may be quite nice to use here~\cite{feng2017learning}, but it does make explicit assumptions in requiring the densities to be differentiable. 
  \item  For an overview of all variational methods, including the use of different divergence metrics see here ~\cite{8588399}
\end{itemize}

\subsection{Using a simple Neural Network regressor}

As a first approach, we use a regressor to learn a function to map
between the input / output pairs that are outputted from the rejection
sampling blocks with in the simulator.

\paragraph{Architecture} For each rejection sampling block we learn a proposal $q(z_1, \ldots, z_i | \eta)$
\subsection{Using variants of VAEs}

@@

\bibliography{refs}
\bibliographystyle{plain}
\appendix

\section{The unconditional acceptance probability}

Given a collection of i.i.d random variables $\{X_{i}\}_{i = 1:n}$ with 
cumulative density function (CDF) $F_{X}$. Then by the Glivenko-Cantelli theorem~\cite{tucker1959generalization} 
we can define the empirical distribution function for $x \in \mathbb{R}$ as:
\begin{equation}
  F_{n}(x) = \frac{1}{n}\sum^{n}_{i=1} \mathbb{I}_{[X_{i}, \infty)}(x)
\end{equation}
And for any fixed $x \in \mathbb{R}$, the Strong law of large numbers ensures that 
$F_{n}(x) \rightarrow F_{X}(x)$ as $n \rightarrow \infty$ as 
\begin{equation}
\mathbb{E}[\mathbb{I}_{[X_{i},\infty)}(x)] = 
 P[\mathbb{I}_{X_{i},\infty)}(x) = 1] = P[X_{i}\leq x] = F_(X)(x)
\end{equation}
where $\mathbb{I}_{A}(x)$  is the indicator function for a set $A$.

\begin{align}
  \text{Prob}(U \leq \frac{p_{X}(z)}{Mg_{Z}(z)} = \mathbb{E}_()
\end{align}
\section{Law of total expectations}

If $X$ is a random variable whose expected value is $\mathbb{E}[X]$ and $Z$ is a random variable on the same probability space
then: \begin{equation}
  \label{eq:lawoftotalexp}
  \mathbb{E}[X] = \mathbb{E}_{p(z)}[\mathbb{E}_{p(x|z)}[X | Z]]
\end{equation}
That is, the expected value of $X$ given $Z$ is the same as the expectation of $X$.
